---
title: "Microarray Data Rspondin KO in Stomach Corpus - GSEA analysis on DGE results"
author: "Hilmar Berger"
output: 
 html_document:
    toc: true
    toc_depth: 3
    toc_float: false
    number_sections: true
    code_folding: hide
    
pdf_document:
    fig_caption: true
date: "`r format(Sys.time(), '%d %B, %Y %H:%M:%S')`"
---

```{r, warning=FALSE, message=FALSE}
rm(list=ls())

suppressMessages(library(xlsx))
library(reshape2)
library(pheatmap)
library(ggplot2)
library(knitr)
library(scales)
library(fgsea)
library(data.table)
library(BiocParallel)

source("./fGSEA_plot_functions.R")

result_folder = "../../../Results/"
set.seed(12345)
```

# Overview

We here run GSEA on DGE results from the 44k arrays. For analysis, a preranked analysis using the R package fGSEA is used that should give similar results to preranked analysis in standard GSEA. 

# MSigDB gene sets

Since MSigDB gene sets use human gene symbols to map genes to pathways we will translate mouse symbols to homologous human symbols using HomologeneDB from NCBI (build 68). 


```{r}
load(file.path(result_folder, "DGE_results_Corpus.Rdata"))
```

```{r}
gsea_gmt_folder = "../../../Data/External/GeneSets/MSigDB_v6.2"
gsea_gmt_files = c("h.all.v6.2.symbols.gmt","c2.all.v6.2.symbols.gmt","c3.all.v6.2.symbols.gmt","c5.bp.v6.2.symbols.gmt","c6.all.v6.2.symbols.gmt","c7.all.v6.2.symbols.gmt")
names(gsea_gmt_files) = c("HALLMARK","PATHWAYS","MOTIF","GO_BP","ONCOGENIC","IMMUNOLOGIC")
gset_collections_human = list()
for (n in names(gsea_gmt_files)) {
  gset_collections_human[[n]] = gmtPathways(paste(gsea_gmt_folder, gsea_gmt_files[[n]], sep="/") )
}
```

We use gene sets from MSigDB.

```{r}
gsea_gmt_files
```

# Other gene sets 

  - Munoz et al Small Intestinal Stem Cell Signature (PMID: 22692129, DOI: 10.1038/emboj.2012.166, Supp Table S3 )  
  - Stange et al Troy+ Chief Cell Signature (Cell 2013, http://dx.doi.org/10.1016/j.cell.2013.09.008, Supp Table S1)


```{r}
known_genes = unique(all_results[[1]]$GeneSymbol)

custom_gene_sets_mouse = list()
intestinal_sc_sets = gmtPathways("../../../Data/External/GeneSets/SC_and_CryptCell_Signatures_mouse.gmt" )
custom_gene_sets_mouse[["Lgr5_SC_Signature_Munoz"]] = intestinal_sc_sets[["Lgr5_SC_Signature_Munoz"]]

tmp = read.table("../../../Data/External/GeneSets/antral lgr5 signature from barker 2010.txt",sep="\t", header=T)
custom_gene_sets_mouse[["Barker_Antral_SC"]] = tmp$Gene.Name

tmp  = read.table("../../../Data/External/GeneSets/Stange_et_al_2013_Troy_signature_supp_table_1.txt", sep="\t", header=T, stringsAsFactors=F)
custom_gene_sets_mouse[["Stange_Troy_signature"]] = unique(tmp$HUGO)

gset_collections_mouse = list(CUSTOM_MOUSE=custom_gene_sets_mouse)

```


## fGSEA analysis on MSigDB

```{r}
homologene_human_mouse = fread("../../../Data/External/homologene_9606_10090.txt")
all_entrez_ids = data.frame(EntrezID=unique(all_results[[1]]$EntrezID))
all_symbols = data.frame(GeneSymbol=unique(all_results[[1]]$GeneSymbol))
a1 = merge(all_symbols, homologene_human_mouse[,c("entrez_1","symbol_1","symbol_2"), with=F], by.x="GeneSymbol", by.y="symbol_2", all.x=T, sort=F)

mouse_to_human=a1
mouse_to_human = subset(mouse_to_human, !is.na(GeneSymbol))
rownames(mouse_to_human) = mouse_to_human$GeneSymbol

```


### Human and Mouse gene sets

For analyzing the human MsigDB gene sets we restrict the DGE data to those probesets that have a homologous gene in mouse and human. For those probesets we select the one with the highest t-score and rank the resulting list which is subsequently used for fGSEA analysis. 

For mouse gene sets (i.e. those derived from mouse experiments) we use the full set of genes in the DGE results after collapsing t-scores by gene and ranking. 

```{r}
data_generation_date = "2019-03-27"
output_folder = result_folder
GSEA_datafile = paste(output_folder,"fGSEA_Results_Corpus.Rdata",sep="/")
if (!file.exists(output_folder)) dir.create(output_folder, recursive=T)

rerun_GSEA = TRUE

if(rerun_GSEA) {
  all_gsea_results = list()
  n_perm = 5000
  n_proc = 8
  bpparam = MulticoreParam(worker=n_proc, manager.hostname="127.0.0.1")
  for (n in names(all_results)) {
    print(paste("Running FGSEA for comparison",n))
    gsea_result_tables = list()
    tmp = subset(all_results[[n]], !is.na(GeneSymbol)) # those are mostly control probes
    
    tmp2 = merge(tmp, mouse_to_human, by="GeneSymbol")
    tmp2 = subset(tmp2, !is.na(symbol_1))
    ranks = unlist(tapply(tmp2$t, tmp2$symbol_1, function(x) x[which(abs(x)==max(abs(x)))] ))
  
    for (gs in names(gset_collections_human)) {
      print(gs)
      fgseaRes <- fgsea(gset_collections_human[[gs]], ranks, minSize=15, maxSize=2000, nperm=n_perm, BPPARAM=bpparam)
      fgseaRes$GeneSetCollection = gs
      gsea_result_tables[[gs]] = fgseaRes
    }
    
    results_human = do.call(rbind, gsea_result_tables)
    
    gsea_result_tables_mouse = list()
    ranks_mouse = unlist(tapply(tmp$t, tmp$GeneSymbol, function(x) x[which(abs(x)==max(abs(x)))] ))
    
    for (gs in names(gset_collections_mouse)) {
      print(gs)
      fgseaRes <- fgsea(gset_collections_mouse[[gs]], ranks_mouse, minSize=15, maxSize=2000, nperm=n_perm, BPPARAM=bpparam)
      fgseaRes$GeneSetCollection = gs
      gsea_result_tables_mouse[[gs]] = fgseaRes
    }
  
    results_mouse = do.call(rbind, gsea_result_tables_mouse)
    results_all = rbind(results_human, results_mouse)
    results_all$FDR_global = p.adjust(results_all$pval, "BH")
    
    all_gsea_results[[n]] = results_all
  }
  save(all_gsea_results, file=GSEA_datafile)
  save(gset_collections_human, gset_collections_mouse, file=file.path(output_folder, "Geneset_collections_Corpus.Rdata"))
} else {
  load(GSEA_datafile)
}
```

Result files are written to folder `r result_folder`

We export all gene sets with less than 15% FDR after global adjustment of p-values per sample. 

```{r}
FDR_threshold = 0.5
if(rerun_GSEA) {
  output_file_prefix = paste(result_folder,"fGSEA_Results_Corpus_", sep="/")
  for (n in names(all_gsea_results)) {
    filename = paste(output_file_prefix, n, ".txt", sep="" )
    tmp = all_gsea_results[[n]]
    tmp$direction= ifelse(tmp$NES > 0, "Up","Down")
    tmp = tmp[order(tmp$NES, decreasing=T),]
    tmp$leadingEdge = unlist(lapply(tmp$leadingEdge, function(x) paste(x, collapse=",")))
    tmp = subset(tmp, FDR_global < FDR_threshold)
    col_order = c("pathway","GeneSetCollection","direction","pval","padj","ES","NES","nMoreExtreme","size","leadingEdge", "FDR_global")
    write.table(tmp[,col_order, with=F], file= filename, row.names=F , sep="\t", dec=".", quote=F, na="")
  }
  
  
  # DGE_Excel_filename = paste(result_folder,"fGSEA_Results",sep="/")
  # txt2xlsx_command = paste("python /data_genome1/SharedSoftware/inhouse_development/tools/txt2xlsx.py -o",DGE_Excel_filename," -L", paste(names(all_gsea_results),collapse=","),  paste(paste('"', output_file_prefix, names(all_gsea_results), '.txt"', sep="" ), collapse=" "), sep=" ")
  # system(txt2xlsx_command)

}
```



```{r, fig.width=12, fig.height=8}
gsea_summary = list()
for (n in names(all_gsea_results)) {
  tmp = subset(all_gsea_results[[n]], padj < 0.05)
  r = as.data.frame(table(tmp$GeneSetCollection, ifelse(tmp$ES<0, "Down","Up")))
  colnames(r) = c("Collection","Direction","Freq")
  r$Comparison = n
  gsea_summary[[n]] = r
}

gsea_summary_df = do.call(rbind, gsea_summary)

ggplot(gsea_summary_df, aes(x=Comparison, y = Freq, fill=Collection)) + geom_bar(stat="identity") + theme_grey(base_size = 14) + theme(axis.text.x = element_text(angle = 60, hjust = 1)) + ggtitle("Number of significant (FDR<5%) gene sets per comparison")

ggplot(gsea_summary_df, aes(x=Comparison, y = Freq, fill=Collection)) + geom_bar(stat="identity") + theme_grey(base_size = 14) + theme(axis.text.x = element_text(angle = 60, hjust = 1)) + ggtitle("Number of significant (FDR<5%) gene sets per comparison") + facet_wrap(~Direction)
```


```{r}
sessionInfo()
```


