---
title: "Microarray Data Rspondin KO in Antrum - DGE Reanalysis with LIMMA"
author: "Hilmar Berger"
output: 
 html_document:
    toc: true
    toc_depth: 3
    toc_float: false
    number_sections: true
    code_folding: hide
    
pdf_document:
    fig_caption: true
date: "`r format(Sys.time(), '%d %B, %Y %H:%M:%S')`"
---

```{r, echo=FALSE, results='hide', messages=FALSE}
rm(list=ls())
suppressMessages(library(limma))
suppressMessages(library(xlsx))
suppressMessages(library(pheatmap))
library(xtable)
library(reshape2)

load("../../../Data/Processed/Rspo_KO_Antrum_micro_array_preprocessed_data.Rdata")
```

# Introduction

This is data from a mouse experiment (Rspo wt vs KO) hybridized as dual channel to Agilent 014568 micro array. 

# MDS

```{r, echo=FALSE, fig.width=8, fig.height=8}

################################################################################
## MDS on normalized data 
cp = palette(rainbow(11))
data_inp = t(as.matrix(MA.n))

d <- dist(data_inp) # euclidean distances between the rows
fit <- cmdscale(d,eig=TRUE, k=2) # k is the number of dim

# plot solution
x <- fit$points[,1]
y <- fit$points[,2]
cc = cp[as.numeric(factor(ed[rownames(data_inp),]$Condition))]
plot(x, y, xlab="Coordinate 1", ylab="Coordinate 2", main="Metric MDS, all samples", type="n", ylim=c(min(y)-50, max(y)+50), xlim=c(min(x)-50, max(x)+200))
points(x, y, col=cc)
#text(x,y,labels=ed[rownames(data_inp),]$sample_ID, col=cp[as.numeric(factor(ed[rownames(data_inp),]$Tissue.Status))])
text(x,y,labels=ed[rownames(data_inp),]$Short, col=cc[as.numeric(as.factor(ed[rownames(data_inp),]$Condition))], cex=.8, pos=4)
################################################################################
```

# DGE

We use separate channel analysis for comparison of KO to WT samples ( Smyth, G.K. and Altman, N.S. (2013).  Separate-channel analysis of two-channel microarrays:recovering inter-spot information.BMC Bioinformatics14, 165).

```{r, echo=FALSE, fig.width=8, fig.height=8, warning=FALSE}
all_results = list()

# Single Channel analysis - all arrays
#sel_arrays = rownames(ed)[ed_014850$chip_barcode %in% c("221485074844_1_1","221485074844_1_2")]
sel_arrays = rownames(ed)
targets2 = targetsA2C(ed[sel_arrays,])
u <- unique(make.names(targets2$Target))
f <- factor(make.names(targets2$Target), levels=u)
design <- model.matrix(~0+f)
colnames(design) <- u
data = MA.n[,sel_arrays]
corfit <- intraspotCorrelation(data, design)
fit <- lmscFit(data, design, correlation=corfit$consensus)

##################################
# NI conditions
##################################
cont.matrix <- makeContrasts("ko - wt",levels=design)
fit2 <- contrasts.fit(fit, cont.matrix)
fit2 <- eBayes(fit2)
all_results[["KO_vs_WT"]] = topTable(fit2, adjust="BH",number=nrow(fit2))


```

## Volcano plot


```{r, volcano, echo=FALSE, fig.width=8, fig.height=8}
all_target_conditions = names(all_results)
par(mfrow=c(1,1), mar=c(4,4,4,2))

for (tc in all_target_conditions) {
  r = all_results[[tc]]
  plot(r$logFC, -log10(r$adj.P.Val),xlab="log2 Fold Change",ylab="-log10(adj. p-val)", ylim=c(0,max(2,max(-log10(r$adj.P.Val),na.rm=T))))
  title(main=tc, sub=paste("(",nrow(subset(r, adj.P.Val < 0.05))," signif. DE genes)",sep="") )
  abline(h=-log10(0.05),col="red")
  abline(v=c(-1,1))
}
```

# Data export 

```{r}
result_folder = "../../../Results/"
```


All result files are exported to `r result_folder`.


```{r, write_tabs, echo=FALSE}
output_file_prefix = paste(result_folder,"Differential_expression_results_Antrum", sep="/")
selected_cols = c("ProbeName", "GeneSymbol", "GeneName","EntrezID","logFC","AveExpr","t","P.Value","adj.P.Val" )
for (tc in all_target_conditions) {
  write.table(all_results[[tc]][,selected_cols], file=paste(output_file_prefix, tc, ".txt", sep="" ), row.names=F , sep="\t", dec=".", quote=F)
}

```

```{r}
output_file = file.path(result_folder, "DGE_Results_Antrum.Rdata")
save(all_results, ed, file=output_file)
```

```{r}
sessionInfo()
```
